---
title: "Bike_analysis"
author: "Hiba"
date: "2023-11-15"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This is the capstone project from the Google Data Analytics certificate. The following six phases of data analysis will guide the analysis:

1. Ask
2. Prepare
3. Process
4. Analyze
5. Share
6. Act

## Ask

#### About the company 

Cyclistic stands out as a bike-share program with over 5,800 bicycles and 600 docking stations. They have two kinds of users: casual and annual members. The director of marketing believes the company’s future success depends on maximizing the number of annual memberships.

#### Key stakeholders

1. Lily Moreno: Director of marketing
2. The Cyclistic executive team
3. The Cyclistic marketing analytics team

#### Business task
Analyze the bike sharing company  Cyclistic's data from January 2023 to September 2023 to determine how annual members and casual riders use Cyclistic bikes differently.

## Prepare

#### Description of the data sources used
The data has been made available online by Motivate International Inc. under this
[license](https://divvybikes.com/data-license-agreement). Data-privacy issues prohibit us from using riders’ personally identifiable information, therefore, it has been removed. The data is current, comprehensive, reliable, and cited.

I have used the data from January 2023 to September 2023.

Download the dataset [here](https://divvy-tripdata.s3.amazonaws.com/index.html).

## Process

#### Import the data
```{r}
library(tidyverse)
files <- list.files(path = "bike_data/")
path_files <- paste("bike_data/",files, sep="")
df <- read_csv(path_files)
summary(df)
```

#### Clean and manipulate the data

The data has been cleaned in the following ways:
1. Missing values are dropped.
2. Duplicates are removed.
3. Rides with duration less than a minute or more than a day will are removed.

The data has been manipulated in the following ways:
1. A new column called ride_length to calculate the ride duration in seconds has been added.
2. A column called start_day that contains the day the ride started has been added.
3. A column called start_month that contains the month the ride started has been added

```{r}
library(tidyr)
library(lubridate)
# remove missing values and duplicates
df_clean <- df %>% drop_na() %>% distinct()
```

```{r}
# add a column called ride_length to calculate the ride duration in seconds
# add a column start_day that contains the day the ride started e,g., Sun
# add a column called start_month that contains the month the ride started e.g., Jan
df_clean <- df_clean %>% mutate(ride_length = ended_at - started_at, start_day = wday(as.Date(started_at), label=TRUE), start_month = month(as.Date(started_at), label=TRUE))
head(df_clean)
```

```{r}
# check min and max of ride_duration as a range check
min(df_clean$ride_length)
max(df_clean$ride_length)
```

```{r}
# remove instances where the duration is 0 or below
df_clean <- df_clean %>% filter(ride_length > 0)
# remove instances where duration is over 24*60*60
df_clean <- df_clean %>% filter(ride_length <= (24*60*60))
```

## Analyze

#### Determine maximum trip duration for each user type
```{r}
df_max_duration <- df_clean %>% group_by(member_casual) %>% summarize(max_trip_len = max(ride_length))
df_max_duration
```

#### Find minimum trip duration for each user type
```{r}
df_min_duration <- df_clean %>% group_by(member_casual) %>% summarize(min_trip_len = min(ride_length))
df_min_duration
```

#### Determine mean trip duration for each user type
```{r}
df_mean_duration <- df_clean %>% group_by(member_casual) %>% summarize(mean_trip_len = mean(ride_length))
df_mean_duration
```

#### Find the day with most bike rides for each user type
```{r}
df_mode_day <- df_clean %>% group_by(member_casual) %>% count(start_day, sort = TRUE) %>% summarize(mode_day = start_day[1])
df_mode_day
```

#### Determine the month with most bike rides for each user type
```{r}
df_mode_month <- df_clean %>% group_by(member_casual) %>% count(start_month, sort = TRUE) %>% summarize(mode_month = start_month[1])
df_mode_month
```

#### Find the count of rides for each day for each user type
```{r}
df_count_day <- df_clean %>% group_by(member_casual, start_day) %>% count(start_day)
df_count_day <- df_count_day %>% rename(count_of_trips = n)
df_count_day
```
#### Find the average ride duration for each day for each user type
```{r}
df_dur_day <- df_clean %>% group_by(member_casual, start_day) %>% summarize(mean_trip_len = mean(ride_length), .groups = "drop")
df_dur_day
```

#### Find the count of rides for each month for each user type
```{r}
df_count_month <- df_clean %>% group_by(member_casual, start_month) %>% count(start_month)
df_count_month <- df_count_month %>% rename(count_of_trips = n)
df_count_month
```

#### Determine the count of type of bikes for each user type
```{r}
df_count_type <- df_clean %>% group_by(member_casual) %>% count(rideable_type)
df_count_type <- df_count_type %>% rename(count_of_trips = n)
df_count_type
```

#### Determine the most popular start stations for casual users
```{r}
df_casual_start_stations <- df_clean %>% filter(member_casual == "casual") %>% count(start_station_name, sort = TRUE) %>% top_n(5)
df_casual_start_stations <- df_casual_start_stations %>% rename(count_of_trips = n)
df_casual_start_stations
```

#### Find the most popular start stations for members
```{r}
df_member_start_stations <- df_clean %>% filter(member_casual == "member") %>% count(start_station_name, sort = TRUE) %>% top_n(5)
df_member_start_stations <- df_member_start_stations %>% rename(count_of_trips = n)
df_member_start_stations
```

#### Find the most popular end stations for casual users
```{r}
df_casual_end_stations <- df_clean %>% filter(member_casual == "casual") %>% count(end_station_name, sort = TRUE) %>% top_n(5)
df_casual_end_stations <- df_casual_end_stations %>% rename(count_of_trips = n)
print(df_casual_end_stations)
# creating auxiliary df for visualization
x <- df_casual_start_stations
x <- x %>% add_column(type = "start")
x <- x %>% rename(station = start_station_name)
y <- df_casual_end_stations
y <- y %>% add_column(type = "end")
y <- y %>% rename(station = end_station_name)
df_casual_stations <- rbind(x, y)
```

#### Determine the most popular end stations for members
```{r}
df_member_end_stations <- df_clean %>% filter(member_casual == "member") %>% count(end_station_name, sort = TRUE) %>% top_n(5)
df_member_end_stations <- df_member_end_stations %>% rename(count_of_trips = n)
df_member_end_stations
```

## Share
}
```{r}
str(df_dur_day)
options(scipen = 999)
df_dur_day$mean_trip_len <- as.numeric(df_dur_day$mean_trip_len)
df_dur_day$mean_trip_len <- round(df_dur_day$mean_trip_len, 2)
ggplot(data = df_dur_day, aes(x = start_day, y = mean_trip_len, fill = member_casual, label = mean_trip_len)) +
  geom_bar(stat = "identity") +
  geom_text(size = 3, position = position_stack(vjust = 0.5)) +
  labs(title = "Average duration of bike trips per day for casual users and annual members", x = "Start day", y = "Average trip length", fill = "user type")
```
* On average, both casual users and annual members use bikes for longer duration on weekends.

```{r}
options(scipen = 999)

ggplot(data = df_count_day, aes(x = start_day, y = count_of_trips, fill = member_casual, label = count_of_trips)) +
  geom_bar(stat = "identity") +
  geom_text(size = 3, position = position_stack(vjust = 0.5)) +
  labs(title = "Count of bike trips per day for casual users and annual members", x = "Start day", y = "Count of trips", fill = "user type")
```

* Casual users use bikes more on weekends indicating that they likely rent bikes for leisure activities
* Annual members tend to rent bikes more on weekdays indicating a more work-related usage.

```{r}
ggplot(data = df_count_month, aes(x = start_month, y = count_of_trips, fill = member_casual, label = count_of_trips)) +
  geom_bar(stat = "identity") +
  geom_text(size = 3, position = position_stack(vjust = 0.5)) +
  labs(title = "Count of bike trips per month for casual users and annual members", x = "Start month", y = "Count of trips", fill = "user type")
```

* The month with most rides for casual users in July.
* The month with most rides for annual members in August.

```{r}
ggplot(data = df_casual_stations, aes(x = station, y = count_of_trips, fill = type, label = count_of_trips)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  geom_text(size = 3, position = position_stack(vjust = 0.5)) +
  labs(title = "The top 5 start and end stations for casual users", x = "Station", y = "Count of trips", fill = "type")

```

The top 5 start stations for casual users is same as their top 5 end stations.

## Act

The following recommendations to convert casual users into annual members are based on the above analysis:

1. **Weekend Conversion Campaign**: Launch a targeted weekend conversion campaign that offers special incentives to casual users who switch to annual memberships. Promote the benefits of having year-round access to bikes on weekends, such as leisurely rides and family outings.

2. **July Promotion**: Leverage the fact that casual users are most active in July. Run a limited-time promotion during this peak month, offering a significant discount on annual memberships, enticing casual users to commit to an annual plan during the period when they use bikes the most.

3. **Top 5 Stations Targeting**: Concentrate marketing efforts on the top 5 stations where casual users frequently start and end their rides. Place eye-catching advertisements and offer on-site assistance to guide these users in the membership conversion process.

By aligning Cyclistic's marketing strategies with these insights, the company can potentially boost our annual membership numbers and enhance overall customer retention.
